---
title: "Google Data Analytics Capstone"
author: "Huma Perveen"
date: "1/3/2022"
output: html_document
---

## Case Study: How Does a Bike-Share Navigate Speedy Success?

### Scenario

I am a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, my team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, my team will design a new marketing strategy to convert casual riders into annual members.

### About the company

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

### Ask

#### Guiding questions: 
How do annual members and casual riders use Cyclistic bikes differently?

#### Business tasks: 
Design marketing strategies aimed at converting casual riders into annual members.

#### Key stakeholders:

* **Lily Moreno:** The director of marketing and your manager. Moreno is responsible for the development of campaigns and initiatives to promote the bike-share program. These may include email, social media, and other channels.
* **Cyclistic marketing analytics team:** A team of data analysts who are responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy.

##### I will produce a report with the following deliverables:
 1. A clear statement of the business task
 2. A description of all data sources used
 3. Documentation of any cleaning or manipulation of data
 4. A summary of analysis
 5. Supporting visualizations and key findings
 6. Top three recommendations based on your analysis

### Prepare

I will use Cyclistic’s historical trip data from Dec2020- Nov2021 to analyze and identify trends.For the purposes of this case study,
[Download the previous 12 months of Cyclistic trip data here.](https://divvy-tripdata.s3.amazonaws.com/index.html) the datasets are appropriate and will enable me to answer the business questions. The data has been made available by Motivate International Inc. under this [license.](https://ride.divvybikes.com/data-license-agreement) 

* This is public data that I will use to explore how different customer types are using Cyclistic bikes. 
* Data is example of ROCCC because it is original data from a reliable organization, comprehensive, current, and cited.
* I downloaded and stored data appropriately. I follow the file naming convention appropriately too.

#### Step 1: Install and Load required packages

```{r}
# for data manipulation, exploration, and visualization.
library(tidyverse)
# for date functions
library(lubridate)
library(scales)
```

#### Step 2: Import Data

12 months trip data from Dec,2020 to Nov,2021

```{r}
tripdata_2020_12 <- read_csv("../source/202012-divvy-tripdata.csv")
tripdata_2021_01 <- read_csv("../source/202101-divvy-tripdata.csv")
tripdata_2021_02 <- read_csv("../source/202102-divvy-tripdata.csv")
tripdata_2021_03 <- read_csv("../source/202103-divvy-tripdata.csv")
tripdata_2021_04 <- read_csv("../source/202104-divvy-tripdata.csv")
tripdata_2021_05 <- read_csv("../source/202105-divvy-tripdata.csv")
tripdata_2021_06 <- read_csv("../source/202106-divvy-tripdata.csv")
tripdata_2021_07 <- read_csv("../source/202107-divvy-tripdata.csv")
tripdata_2021_08 <- read_csv("../source/202108-divvy-tripdata.csv")
tripdata_2021_09 <- read_csv("../source/202109-divvy-tripdata.csv")
tripdata_2021_10 <- read_csv("../source/202110-divvy-tripdata.csv")
tripdata_2021_11 <- read_csv("../source/202111-divvy-tripdata.csv")
```

Check column name should be consistent

```{r}
colnames(tripdata_2020_12)
colnames(tripdata_2021_01)
colnames(tripdata_2021_02)
colnames(tripdata_2021_03)
colnames(tripdata_2021_04)
colnames(tripdata_2021_05)
colnames(tripdata_2021_06)
colnames(tripdata_2021_07)
colnames(tripdata_2021_08)
colnames(tripdata_2021_09)
colnames(tripdata_2021_10)
colnames(tripdata_2021_11)
```

Check the structure of dataframes for any incongruencies

```{r}
str(tripdata_2020_12)
str(tripdata_2021_01)
str(tripdata_2021_02)
str(tripdata_2021_03)
str(tripdata_2021_04)
str(tripdata_2021_05)
str(tripdata_2021_06)
str(tripdata_2021_07)
str(tripdata_2021_08)
str(tripdata_2021_09)
str(tripdata_2021_10)
str(tripdata_2021_11)
```

Combine all dataframes into one dataframe for one year

```{r}
all_tripdata <- bind_rows(tripdata_2020_12, tripdata_2021_01, tripdata_2021_02, tripdata_2021_03, 
                          tripdata_2021_04, tripdata_2021_05, tripdata_2021_06, tripdata_2021_07, 
                          tripdata_2021_08, tripdata_2021_09, tripdata_2021_10, tripdata_2021_11)
```

Remove start_lat, start_lng, end_lat and end_lng columns because I dont need these feature in my data analysis.

```{r}
all_trips <- all_tripdata %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

### Process

Data cleaning and add data for analysis.

Due to the large dataset I will use R for data cleaning and analysis.

```{r}
# list all column names
colnames(all_trips) 
```

```{r}
# dimensions of data frame
dim(all_trips)  
```

```{r}
# number of rows in data frame
nrow(all_trips) 
```

```{r}
# Only show first 6 rows of data frame
head(all_trips)
```

```{r}
# only last 6 rows of data frames
tail(all_trips)
```

```{r}
# Brief structure of data frame with column name and data type
str(all_trips)
```

```{r}
# summary of data frame, mainly numerics
summary(all_trips)
```

Now I need to add column for date, month, day, year, and day of week for each ride  to aggregate

```{r}
# ride data for each day, month and year

all_trips$date <- as.Date(all_trips$started_at) # default date format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%B")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$hour <- format(as_datetime(all_trips$started_at), "%H")
```

```{r}
# confirm that column has been added
colnames(all_trips)
```

Now I add ride_length to all_trips data frame to calculate ride length in seconds for each trip

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

I inspect the data frame again to check column is added or not and what is data type of ride_length

```{r}
str(all_trips)
```

```{r}
# Convert "ride_length" to numeric so we can run calculations on the data
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

# check is it numeric?
is.numeric(all_trips$ride_length)
```

```{r}
# Convert "hour" to numeric
all_trips$hour <- as.numeric(as.character(all_trips$hour))

# check is it numeric?
is.numeric(all_trips$hour)
```

I check the data frame again

```{r}
str(all_trips)
```

#### Remove bad data 

I remove data where ride length is 0 or less than 0 and when bikes were taken out of docks for quality check, and check the number of rows remaining


```{r}
all_trips_v2 <- all_trips[!(all_trips$rideable_type =="docked_bike" | all_trips$ride_length <= 0),]
nrow(all_trips_v2)
```

#### Remove NA from new data frame

check the summary of data frame

```{r}
all_trips_v3 <- drop_na(all_trips_v2)

dim(all_trips_v3) 
View(all_trips_v3)
summary(all_trips_v3)
```

### Analyze

#### Descriptive analysis on ride_length (seconds)

```{r}
mean(all_trips_v3$ride_length) #average ride length (total ride length / rides)
median(all_trips_v3$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v3$ride_length) #longest ride
min(all_trips_v3$ride_length) #shortest ride
```

Compare ride length by members and casual users

```{r}
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual, FUN = mean)
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual, FUN = median)
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual, FUN = max)
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual, FUN = min)
```

Assign the correct order to each day of the week

```{r}
all_trips_v3$day_of_week <- 
  ordered(all_trips_v3$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

See the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual + all_trips_v3$day_of_week, FUN = mean)
```

Now find number of rides per day for members and casual users

```{r}
all_trips_v3 %>%
  group_by(member_casual, day_of_week) %>%
  summarize(num_of_rides = n(), .groups = "drop") %>%
  arrange(day_of_week)
```

Analyze ridership data by riders type and weekday

```{r}
all_trips_v3 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)	
```

Assign the correct order to each month of the year

```{r}
all_trips_v3$month <-
  ordered(all_trips_v3$month, levels = c("December", "January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November"))
```

Find monthly rides by members and casual users

```{r}
all_trips_v3 %>%
  group_by(member_casual, month) %>%
  summarise(number_of_ride = n(), .groups = 'drop') %>%
  arrange(month)
```

Analyze ridership data by riders type and month

```{r}
all_trips_v3 %>%
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, month)	
```

Average ride time by each month for members vs casual users

```{r}
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual + all_trips_v3$month, FUN = mean)
```

Total number of rides by members or casual users

```{r}
all_trips_v3 %>%
  group_by(member_casual) %>%
  summarize(number_of_rides = n() , .groups = 'drop')
```

Hourly analysis for ridership for casual riders

```{r}
all_trips_v4 <- all_trips_v3 %>%
  group_by(member_casual, hour) %>%
  summarize(number_of_rides = n(), average_duration = mean(ride_length), .groups = 'drop') %>%
  filter(member_casual =="casual")
  
max(all_trips_v4$number_of_rides)
min(all_trips_v4$number_of_rides)
mean(all_trips_v4$number_of_rides)
median(all_trips_v4$number_of_rides)
View(all_trips_v4) 
```

Hourly analysis for riders type for member riders

```{r}
all_trips_v5 <- all_trips_v3 %>%
   group_by(member_casual, hour) %>%
   summarize(number_of_rides = n(),average_duration = mean(ride_length), .groups = 'drop') %>%
   filter(member_casual =="member")
 
 max(all_trips_v5$number_of_rides)
 min(all_trips_v5$number_of_rides)
 mean(all_trips_v5$number_of_rides)
 median(all_trips_v5$number_of_rides)
 View(all_trips_v5)
```

### Share

#### Create Visualizations

Visualization of numbers of rides by members and casual users

```{r}
all_trips_v3 %>%
  group_by(member_casual) %>%
  summarize(number_of_rides = n() , .groups = 'drop') %>%
  ggplot(aes(x = member_casual, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Number of Rides by Rider Types", x = "Riders Type", y = "Number of Rides", fill = "Riders Type") +
  scale_y_continuous(labels = scales::comma)
```
  
Visualization of number of rides by days of week for riders type

```{r}
all_trips_v3 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_rides = n() , .groups = 'drop') %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = weekday)) +
  geom_col(position = "dodge") + 
  facet_wrap(~ member_casual) +
  labs(title = "Number of Rides Grouped by Day of Week", subtitle = "Faceted by Rider Types", x = "Days of Week", y = "Number of Rides", fill = "Days of Week") + 
  scale_y_continuous(labels = scales::comma)
```

Visualization of monthly number of rides by riders type

```{r}
all_trips_v3 %>%
  group_by(member_casual, month) %>%
  summarize(number_of_rides = n(), .groups = 'drop') %>%
  arrange(month) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) + 
  geom_bar(position = "dodge", stat = "identity") + 
  theme(axis.text.x = element_text(angle = 65)) +
  labs(title = "Monthly Number of Rides by Members and Casuals", 
       x = "Months", y = "Number of Rides", fill = "Riders Type") +
  scale_y_continuous(labels = scales::comma)
```

Visualization for number of rides by hours of the day for riders type

```{r}
all_trips_v3 %>%
  group_by(member_casual, hour) %>%
  summarize(number_of_rides = n(), .group = "drop") %>%
  ggplot(aes(x = hour, y = number_of_rides, color = member_casual)) +
  geom_line() +
  scale_x_time(breaks = c(0, 3, 6, 9, 12, 15, 18, 21), 
               labels = c("12 AM", "3 AM", "6 AM", "9 AM", "12 PM", "3 PM", "6 PM", "9 PM")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Numbers of Rides by Hours of Day", 
       x = "Hours", y = "Number of Rides", color = "Riders Type") 
```

Visualization for average ride length per day by riders type

```{r}
all_trips_v3 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(average_ride_length = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_ride_length, fill = weekday)) +
  geom_col(position = "dodge") +
  facet_wrap(~ member_casual) +
  labs(title = "Average Ride Length by Days of Week", subtitle = "Faceted by Riders Type",
       x = "Days of Week", y = "Average Ride Length(Seconds)", fill = "Days of Week") +
  scale_y_continuous(labels = scales::comma)
```

Visualization for average ride length per month by riders type

```{r}
all_trips_v3 %>%
  group_by(member_casual, month) %>%
  summarize(average_ride_length = mean(ride_length)) %>%
  arrange(member_casual, month) %>%
  ggplot(aes(x = month, y = average_ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title = "Average Ride Length From Dec,2020 To Nov,2021", x = "Months", 
       y = "Average Ride Length(Seconds)", fill = "Riders Type") +
  scale_y_continuous(labels = scales::comma)
```

Visualization for average ride length per hour by riders type

```{r}
all_trips_v3 %>%
  group_by(member_casual, hour) %>%
  summarize(average_ride_length = mean(ride_length)) %>%
  ggplot(aes(x = hour, y = average_ride_length, color = member_casual)) +
  geom_line() +
  scale_x_time(breaks = c(0, 3, 6, 9, 12, 15, 18, 21), 
               labels = c("12 AM", "3 AM", "6 AM", "9 AM", "12 PM", "3 PM", "6 PM", "9 PM")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average Ride Length by Hours of Day",
       x = "Hours", y = "Average Ride Length(Seconds)", color = "Riders Type")
```

### Analysis Findings

* From my analysis, it's clear that casual riders have less number of rides than members but their average ride length (duration in seconds) is comparatively more than members.
* From hourly data, we can say that casuals ride length has sharp rise around 7:30 AM till 10:00 AM while membership riders have almost constant average ride length.
* Another interesting finding is that casuals are more active at week end in compare to members who are using more bike rides on weekdays.

### Conclusion

* To attract casual users, short term membership over weekends can be good option.
* Membership discounts can also be planned for summer and autumn months (from May to October).

 
 
 
